---
title: "Intro to Data Frames in R"
subtitle: "Environmental Data Analytics"
author: "Luana Lima"
date: "Fall 2023"
output: pdf_document
---

```{r setup, include=FALSE}
#will not print the chunk 
#echo=Falue -> only print out the result 
knitr::opts_chunk$set(echo = TRUE)
```

## Data frame definition

A **data frame** is used for storing data tables. It a list a vector of equal length.

When we import data to R, data frame is the preferred way for storing the data because columns can have different modes (character, numeric, integer, logical, complex).

## Data frame built-in example

Let's look into a built-in data frame from package "datasets" - cars. The data give the speed of cars and the distances taken to stop.

```{r cars}
#console type data() -> list all availble dataset within R
cars
mtcars 
```

Note that it has 2 columns and 50 rows.

## Data frame columns

Suppose you want just the column speed. How would you access that data?

```{r}
cars$speed
#use $ to specify the column in the dataset 
```


```{r}
mtcars$cyl

```
How would you store it on another object?

```{r}
car_speed <- cars$speed
car_speed
```

## Transforming object in a data frame

Is the new object you create a data frame?

```{r}
#Option 1
class(car_speed)

#Option 2
is.data.frame(car_speed)
#to confirm if car_speed is a data frame 
#is.data.frame(name_dataframe)
```

How could you make it a data frame?

```{r}
df_car_speed <- as.data.frame(car_speed)
#as.data.frame(name_dataframe) -> convert data type to data frame 
#data.frame: assign the type as data frame 
df_car_speed
class(df_car_speed)
```

## Adding columns to a data frame

How could you add columns to *df_car_speed*?

```{r}
car_dist <- cars$dist

#Option 1
df <- cbind(df_car_speed,car_dist)  
#rbind(): similarly rows could be added using rbind()
#cbind(dataframe1, dataframe2): combine data frame by column
#check: combine data frames based on 
class(df)
df

#Or Option 2 - transform into a data frame before binding
df_car_dist <- as.data.frame(car_dist)  #op2
df_opt2 <- cbind(df_car_speed,df_car_dist)
class(df_opt2)
df_opt2

```

Note that when we transformed the vector in a data frame the name of the vector became the column name.

```{r}
colnames(df)  #or simply names()
names(df)
#to check to name of the column 
#name(name_dataframe)
#check: if it is the same for rows names 
```

## Creating a data frame

How would you create a data frame?

```{r}
#useful function data.frame("column_name1"=name_variable1, "column_name2"=name_variable2)
create_df <- data.frame("speed"=car_speed,"dist"=car_dist)
create_df
```

## Data frame functions

Some useful functions to use with data frames.

```{r}
ncol(df)  
nrow(df)  
length(df)  #same as ncol
summary(df)
head(df) #show the first 6 rows of df
```

```{r}
#If you know the number of the column you want you can refer to that to access column
df1<-df[,1]

#df[row, column]
df2<-df[1:10, 1] 
#slice row 1 to 10, show column 1
df3<-df[c(1,3,4),1] 
#only retrieve row #1,#3, #4 (in order) with the first column
df4<- df[c(1,3,4),] 
#only retrieve row #1,#3, #4 (in order) with all column
df5<-df[c(1,3,4),]$car_speed
# if the order of the column is unsure
df6<-df$car_speed[c(1,3,4),]
#error 
#since df$car_speed is a vector, will not have the column
#[c(1,3,4),] : this is only applied to the data frame 
#alternative: assign df$car_speed as date frame (as.dataframe(df$car_speed))

#you could also use this notation to delete columns
df <- df[,-2]
df
```
