---
title: "Assignment 4: Data Wrangling"
author: "Ina Liao"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.

The completed exercise is due on Thursday, Sept 28th @ 5:00pm.

## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.
```{r Install Packages}
#install.packages("tidyverse")
#install.packages("lubridate")
#install.packages("here")
#install.packages("dplyr")
```
```{r Loading Packages}
library(tidyverse)
library(lubridate)
library(here)
library(dplyr)
```
1b. Check your working directory.
```{r Working Directory}
getwd()
```

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

```{r Import Files}
df_EPA_O3_2018<-read.csv("/Users/inaliao/Desktop/EDE_Fall2023/Data/Raw/EPA/EPAair_O3_NC2018_raw.csv",stringsAsFactors = TRUE)
df_EPA_O3_2019<-read.csv("/Users/inaliao/Desktop/EDE_Fall2023/Data/Raw/EPA/EPAair_O3_NC2019_raw.csv",stringsAsFactors = TRUE)
df_EPA_PM_2018<-read.csv("/Users/inaliao/Desktop/EDE_Fall2023/Data/Raw/EPA/EPAair_PM25_NC2018_raw.csv",stringsAsFactors = TRUE)
df_EPA_PM_2019<-read.csv("/Users/inaliao/Desktop/EDE_Fall2023/Data/Raw/EPA/EPAair_PM25_NC2019_raw.csv",stringsAsFactors = TRUE)
```

2. Apply the `glimpse()` function to reveal the dimensions, column names, and structure of each dataset. 

```{r setup, message = FALSE}
#1a structure of each data set
glimpse(df_EPA_O3_2018)
glimpse(df_EPA_O3_2019)
glimpse(df_EPA_PM_2018)
glimpse(df_EPA_PM_2019)

#1b dimentions of each data set
dim(df_EPA_O3_2018)
dim(df_EPA_O3_2019)
dim(df_EPA_PM_2018)
dim(df_EPA_PM_2019)

#1c column names of each dataset
colnames(df_EPA_O3_2018)
colnames(df_EPA_O3_2019)
colnames(df_EPA_PM_2018)
colnames(df_EPA_PM_2019)

#2 distribution of values in each column of the dataset.
summary(df_EPA_O3_2018)
summary(df_EPA_O3_2019)
summary(df_EPA_PM_2018)
summary(df_EPA_PM_2019)

```


## Wrangle individual datasets to create processed files.

3. Change the Date columns to be date objects.
```{r Date Formating}
df_EPA_O3_2018$Date<-as.Date(df_EPA_O3_2018$Date,format="%m/%d/%Y")
df_EPA_O3_2019$Date<-as.Date(df_EPA_O3_2019$Date,format="%m/%d/%Y")
df_EPA_PM_2018$Date<-as.Date(df_EPA_PM_2018$Date,format="%m/%d/%Y")
df_EPA_PM_2019$Date<-as.Date(df_EPA_PM_2019$Date,format="%m/%d/%Y")

```


4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
```{r Subset}
#select specific columns and create another subset
df_EPA_O3_2018_subset<-df_EPA_O3_2018 %>%
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
head(df_EPA_O3_2018_subset)

df_EPA_O3_2019_subset<-df_EPA_O3_2019 %>%
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
head(df_EPA_O3_2019_subset)

df_EPA_PM_2018_subset<-df_EPA_PM_2018 %>%
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
head(df_EPA_PM_2018_subset)

df_EPA_PM_2019_subset<-df_EPA_PM_2019 %>%
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
head(df_EPA_PM_2019_subset)
```

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
```{r Replace Cell Values }
PM_replace_AQS<-function(x){
  ifelse(x=="Acceptable PM2.5 AQI & Speciation Mass","PM2.5","PM2.5")
}
df_EPA_PM_2018_subset$AQS_PARAMETER_DESC<-PM_replace_AQS(df_EPA_PM_2018_subset$AQS_PARAMETER_DESC)
df_EPA_PM_2019_subset$AQS_PARAMETER_DESC<-PM_replace_AQS(df_EPA_PM_2019_subset$AQS_PARAMETER_DESC)
head(df_EPA_PM_2018_subset)
head(df_EPA_PM_2019_subset)
```

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r Export Datasets}
#3
write.csv(df_EPA_O3_2018_subset, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_O3_NC2018_Processed.csv", row.names=FALSE)

#4
write.csv(df_EPA_O3_2019_subset, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_O3_NC2019_Processed.csv", row.names=FALSE)

#5
write.csv(df_EPA_PM_2018_subset, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_PM_NC2018_Processed.csv", row.names=FALSE)

#6
write.csv(df_EPA_PM_2019_subset, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_PM_NC2019_Processed.csv", row.names=FALSE)

```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.
```{r Create O3 & PM Data Frames}
#combine data frame based on AIQ parameters
#combine the original data sets rather than subsets 
df_EPA_O3<-rbind(df_EPA_O3_2018,df_EPA_O3_2019)
df_EPA_PM <- rbind(df_EPA_PM_2018,df_EPA_PM_2019)

#check if the column names are identical 
colnames(df_EPA_O3)
colnames(df_EPA_PM)
head(df_EPA_PM);tail(df_EPA_PM)
```
```{r Rename Column Names}
#rename columns names so PM2.5 and O3 data frames could be combined 
colnames(df_EPA_O3)[colnames(df_EPA_O3)=="Daily.Max.8.hour.Ozone.Concentration"]<-"Daily.Max.8.hour Concentration"
colnames(df_EPA_PM)[colnames(df_EPA_PM)=="Daily.Mean.PM2.5.Concentration"]<-"Daily.Max.8.hour Concentration"
colnames(df_EPA_O3)
colnames(df_EPA_PM)

#replace cells values: fill all cells in AQS_PARAMETER_DESC with "PM2.5"
#use previous defined function, PM_replace_AQS
df_EPA_PM$AQS_PARAMETER_DESC<-PM_replace_AQS(df_EPA_PM$AQS_PARAMETER_DESC)
head(df_EPA_PM)
```
```{r Merge O3 and PM}
df_EPA<-rbind(df_EPA_O3,df_EPA_PM)
head(df_EPA);tail(df_EPA)
```

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include only sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.
  
```{r With Pipe}
df_EPA_subset<-df_EPA %>%
  #filter sites names
  filter(Site.Name=="Linville Falls" | Site.Name=="Durham Armory"| Site.Name=="Leggett"| Site.Name=="Hattie Avenue"| Site.Name=="Clemmons Middle"| Site.Name=="Mendenhall School" | Site.Name=="Frying Pan Mountain"| Site.Name=="West Johnston Co."| Site.Name=="Garinger High School"| Site.Name=="Castle Hayne"| Site.Name=="Pitt Agri. Center"| Site.Name=="Bryson City"|Site.Name== "Millbrook School") %>%
  #calculate mean AIQ, mean latitude, and mean longitude
  group_by(Date,Site.Name,AQS_PARAMETER_DESC,COUNTY) %>%
  summarize(Mean_AQI = mean(DAILY_AQI_VALUE), 
            Mean_Latitude = mean(SITE_LATITUDE),
            Mean_Longitude = mean(SITE_LONGITUDE))%>%
  mutate(Month=month(Date),Year=year(Date))
head(df_EPA_subset)
dim(df_EPA_subset)

```

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

```{r Seperate AIQ Values Based on Parameters}
#seperate mean AIQ values based on AIQ parameters 
df_EPA_subset_spread<- df_EPA_subset %>%
  pivot_wider(names_from="AQS_PARAMETER_DESC",values_from="Mean_AQI") 

#rename columns' names 
colnames(df_EPA_subset_spread)[colnames(df_EPA_subset_spread)=="PM2.5"]<-"Mean_AQI_PM2.5"
colnames(df_EPA_subset_spread)[colnames(df_EPA_subset_spread)=="Ozone"]<-"Mean_AQI_O3"
```

```{r Merge Locations }
df_EPA_subset_spread<- df_EPA_subset_spread %>%
  group_by(Date,COUNTY) 
head(df_EPA_subset_spread);tail(df_EPA_subset_spread)
```

10. Call up the dimensions of your new tidy dataset.

```{r Dimensions}
dim(df_EPA_subset_spread)
```
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r}
#7 O3, 2018 & 2019
write.csv(df_EPA_O3, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_O3_NC1819_Processed.csv", row.names=FALSE)

#8 PM, 2018 & 2019
write.csv(df_EPA_PM, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_PM_NC1819_Processed.csv", row.names=FALSE)

#9 O3 & PM, 2018 & 2019
write.csv(df_EPA, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_O3_PM_NC1819_Processed.csv", row.names=FALSE)

#10 group by dates and location, generate daily mean AIQ values
write.csv(df_EPA_subset, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_O3_PM_NC1819_Processed.csv", row.names=FALSE)

#11 Spread AIQ values for ozone and PM2.5 
write.csv(df_EPA_subset_spread, file="/Users/inaliao/Desktop/EDE_Fall2023/Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv", row.names=FALSE)

```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r}

#12
df_EPA_subset_summary <- df_EPA_subset_spread %>%
  group_by(Site.Name, Month,Year) %>%
  summarize(Mean_AQI_PM2.5 = mean(Mean_AQI_PM2.5), 
            Mean_AQI_O3 =mean(Mean_AQI_O3),
            Mean_Latitude = mean(Mean_Latitude),
            Mean_Longitude = mean(Mean_Longitude)) %>%
  drop_na(Mean_AQI_O3)
head(df_EPA_subset_summary)

#13
dim(df_EPA_subset_summary)

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: na.omit will remove rows with any missing values, and thus, we can not keep the rows with missing mean PM2.5 values.
