---
title: "Assignment 8: Time Series Analysis"
author: "Ina Liao"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r Set up, message=FALSE}
library(tidyverse)
library(lubridate)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(here)
library(ggplot2)
library(ggthemes)
here()
```

```{r Plot Theme, message=FALSE}
plot_theme<-theme_calc()+
  theme(
    #plot background 
    plot.background=element_rect(color="gray"),
    
    #plot title 
    plot.title=element_text(color="black",hjust=0.5,vjust=1),
    
    #axis labels 
    axis.text=element_text(color="black"),
    
    #legend 
    legend.key=element_rect(color="white"),
    legend.background=element_rect(color="white"),
    legend.position="right")

theme_set(plot_theme)
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r Import data, message = FALSE}
#1

EPAair_2010<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv"), stringsAsFactors = TRUE)
EPAair_2011<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv"), stringsAsFactors = TRUE)
EPAair_2012<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv"), stringsAsFactors = TRUE)
EPAair_2013<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv"), stringsAsFactors = TRUE)
EPAair_2014<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv"), stringsAsFactors = TRUE)
EPAair_2015<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv"), stringsAsFactors = TRUE)
EPAair_2016<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv"), stringsAsFactors = TRUE)
EPAair_2017<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv"), stringsAsFactors = TRUE)
EPAair_2018<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv"), stringsAsFactors = TRUE)
EPAair_2019<-read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv"), stringsAsFactors = TRUE)
GaringerOzone<-rbind(EPAair_2010,EPAair_2011,EPAair_2012,EPAair_2013,EPAair_2014,EPAair_2015,EPAair_2016,EPAair_2017,EPAair_2018,EPAair_2019)
#GaringerOzone
```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r Data wrangle, message=FALSE}
# 3
GaringerOzone$Date<-mdy(GaringerOzone$Date)

# 4
colnames(GaringerOzone)
GaringerOzone<-GaringerOzone %>%
  select(Date,Daily.Max.8.hour.Ozone.Concentration,DAILY_AQI_VALUE)

# 5
# filling missing days with NA
# create a sequence as a data frame from 2010-01-01 until 2019-12-31
seq_date<-as.character(seq(as.Date("2010-01-01"), as.Date("2019-12-31"), by = "days"))
df_date<-as.data.frame(seq_date)
# rename column name
colnames(df_date)<-"Date"
df_date$Date<-ymd(df_date$Date)

# 6
GaringerOzone<-left_join(df_date,GaringerOzone,by="Date")
# notes: the order of left_join matters

#rename column names 
GaringerOzone<-GaringerOzone%>%
  rename(Ozone=Daily.Max.8.hour.Ozone.Concentration)%>%
  rename(AQI=DAILY_AQI_VALUE)
```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r Visualize, message=FALSE, fig.cap="Ozone Concentrations over Time"}
#7
ggplot(GaringerOzone,aes(x=Date,y=Ozone))+
  geom_line()+
  geom_smooth(method=lm,color="blue")+
  labs(x="",y="ozone concentration (ppm)")
# missing data is removed in this plot 
```

>Answer:The plot does suggest a clear trend in ozone concentration over time. Further testing is needed to determine whether the data follows a trend. 

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r Missing data, message=FALSE, fig.cap="Ozone Concentrations over Time"}
#8
# check missing observations 
summary(GaringerOzone$Ozone)

# add a new column that replace NA by interpolation
GaringerOzone.clear<-GaringerOzone%>%
  mutate(Ozone_clean=zoo::na.approx(Ozone)) 
summary(GaringerOzone$Ozone_clear)

# use the red line to highlight the missing data
ggplot(GaringerOzone.clear) +
  geom_line(aes(x = Date, y = Ozone_clean), color = "red") +
  geom_line(aes(x = Date, y = Ozone), color = "black") +
  labs(x="",y="ozone concentration (ppm)") 
```

> Answer: The ozone concentration data exhibit a relatively smooth and continuous trend over the time; in addition, the data only have a short period of missing data, and thus, linear interpolation may provide a better approximation than piecewise constant and spline interpolation. 

9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r Monthly Data,message=FALSE}
#9
GaringerOzone.monthly <- GaringerOzone.clear %>%
  separate(Date, c("Year","Month","Day"),"-") %>%
  group_by(Year,Month) %>%
  summarize(Ozone_clean=mean(Ozone_clean)) %>%
  mutate(Date=my(paste0(Month,"-",Year))) %>%
  select(Date,Ozone_clean)
```

10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r Time series objects, message=FALSE}
#10

# daily time series 

# specific the first month and year 
f_month_daily<-month(first(GaringerOzone.clear$Date))
f_year_daily<-year(first(GaringerOzone.clear$Date))
# create time series objects 
GaringerOzone.daily.ts<-ts(GaringerOzone.clear$Ozone_clean,start=c(f_year_daily,f_month_daily),frequency=365)


# monthly time series 

# specific the first month and year 
f_month_monthly<-month(first(GaringerOzone.monthly$Date))
f_year_monthly<-year(first(GaringerOzone.monthly$Date))
# create time series objects 
GaringerOzone.monthly.ts<-ts(GaringerOzone.monthly$Ozone_clean,start=c(f_year_monthly,f_month_monthly),frequency=12)
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r Plot, message=FALSE}
#11

# daily time series 
# decompose 
GaringerOzone.daily.decompose<-stl(GaringerOzone.daily.ts,s.window="periodic")
# plot 
plot(GaringerOzone.daily.decompose)


# monthly time series 
# decompose 
GaringerOzone.monthly.decompose<-stl(GaringerOzone.monthly.ts,s.window="periodic")
# plot 
plot(GaringerOzone.monthly.decompose)
```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?


```{r Monotonic trend analysis, message=FALSE}
#12

# run Seasonal Mann-Kendall test
# Ho: data is stationary 
# Ha: data follow a trend

Ozone_trend<- Kendall::SeasonalMannKendall(GaringerOzone.monthly.ts)
Ozone_trend_season<- trend::smk.test(GaringerOzone.monthly.ts)
summary(Ozone_trend) 
# do not reject Ho (p-value>0.01), meaning that the data does not follow a trend 
summary(Ozone_trend_season)
```

> Answer: The plot suggests that the ozone concentration data might have seasonality, so applying seasonal Mann-Kendall to determine if the data is stationary is appropriate. We can also remove the seasonality in the dataset, and apply other tests to analyze the underlying trend over time. 

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r Plot monthly, message=FALSE}
# 13
plot_Ozone<-ggplot(GaringerOzone.monthly,aes(x=Date,y=Ozone_clean))+
  geom_point()+
  geom_line()+
  geom_smooth(method=lm,color="blue")+
  labs(x=" ",y="ozone concentration (ppm)")
print(plot_Ozone)
```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: The ozone concentration data does not have a statistically significant trend over 2010 at the station (p-bvalue>0.01).

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r Other tests}
#15

# remove seasonality
# extract the components of monthly time sereis and turn it into data frame 
colnames(GaringerOzone.monthly.decompose$time.series)
Ozone_monthly_components<-as.data.frame(GaringerOzone.monthly.decompose$time.series[,2:3])

#add date and observed to components data frame 
Ozone_monthly_components<-Ozone_monthly_components%>%
  mutate(Ozone_clean=GaringerOzone.monthly$Ozone_clean,
         Date=GaringerOzone.monthly$Date)

#16

# create another monthly time series 
# specific the first month and year 
f_month_monthly_wo<-month(first(Ozone_monthly_components$Date))
f_year_monthly_wo<-year(first(Ozone_monthly_components$Date))
# create time series objects 
Ozone_monthly_components.ts<-ts(Ozone_monthly_components$Ozone_clean,start=c(f_year_monthly,f_month_monthly),frequency=12)

# Mann-Kendall
Ozone_MannKendall_all<- Kendall::MannKendall(Ozone_monthly_components.ts)
summary(Ozone_MannKendall_all) 
#reject Ho (p-value>0.1)
```

> Answer: The ozone concentration data does not have a statistically significant trend over 2010 at the station (p-bvalue>0.01), consistent with the results of the Seasonal Mann-Kendall test.
