---
title: "A06 - Crafting Reports"
author: "Ina Liao"
date: "Fall 2023"
output: pdf_document
fig_height: 2
fig_width: 3
geometry: margin=2.54cm
toc: True 
---


\listoftables
\listoffigures

## Objectives:

1.  More practice with R code chunk options
2.  Gain proficiency with figures, tables (w/`Kable`) table of contents, etc.
3.  Debugging knitting issues

## Directions

1.  Rename this file `<FirstLast>_A06_CraftingReports.Rmd` (replacing `<FirstLast>` with your first and last name).

2.  Change "Student Name" on line 3 (above) with your name.

3.  Work through the tasks, **creating code and output** that fulfill each instruction.

4.  Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.

5.  Be sure to **answer the questions** in this assignment document.

6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.

7.  **Be sure that you also commit and push your final Rmd document to your GitHub account**.

## Task 1 - Basic Markdown

Using markdown, create a table beneath the `Table: EPA Air Quality` line below that summarizes the metadata of the EPA Air Quality data. The first column should have the header "Item" and should include the the three metadata attribute item names: "Source", "Date", and "Filename". The second column should have the header "Value" and include the metadata values: "EPA Air Quality System (AQS)", "2018-2019", and "EPAair_O3_PM25_NC1819_Processed.csv". The first column should be aligned to the right and the second to the left.

```{r Install Packages, warning=FALSE,error=FALSE,echo=FALSE}
#options(repos = "https://cran.rstudio.com/")
#install.packages("ggplot2")
#install.packages("tidyverse")
#install.packages("here")
#install.packages("dplyr")
#install.packages("shiny")
#install.packages("lubridate")
#install.packages("knitr")
```

```{r Loading Library,message=FALSE,warning=FALSE,echo=FALSE}
library(ggplot2)
library(tidyverse)
library(here)
library(dplyr)
library(shiny)
library(knitr)
```

```{r Create Table, warning=FALSE, error=FALSE,echo=FALSE, fig.cap="Table: EPA Air Quality"}
#create a data frame
df_table<-data.frame(
  item=c("Source","Date","Filename"),
  value=c("EPA Air Quality System (AQS)","2018-2019","EPAair_O3_PM25_NC1819_Processed.csv"))

#create summary table
#name another variable to store the dataframe, or the data frame will be printed out 
table<-df_table%>%
  summarize(ItemName = item,
            Value=value)

#rename column names
colnames(table) <-c("Item Name","Value")

#print out the table
knitr::kable(table,caption="EPA Air Quality")

```

------------------------------------------------------------------------

## Task 2 - Import packages and data, suppressing messages

Set the following R code chunk so that it runs when knit, but no messages, errors, or any output is shown. The code itself, however, should be displayed.

```{r Import Data, warning=FALSE,error=FALSE,echo=TRUE}
#import EPA data (from the processed_KEY folder) & fix dates
epa_data <- read.csv(
  here("data","processed_KEY","EPAair_O3_PM25_NC1819_Processed.csv"),
  stringsAsFactors = TRUE) %>% 
  mutate(Date = ymd(Date))
#note: "echo=TRUE" will only display the output
```

------------------------------------------------------------------------

## Task 3: Creating tables

Set the following R code chunk to display two tables, using knitr's `kable()` function, one listing the mean PM2.5 concentrations for each county, and the other the same except for Ozone. The titles should be "Mean Particulates (2.5mm)" and "Mean Ozone", respectively. And the column names should be "County" and "µg/m3" for both tables.

Customize the chunk options such that the code is run but is not displayed in the knitted document. The output, however, should be displayed.

> **TIPS:**
>
> -   Use `"$\\mu g/m^3$"` as a column name to generate a nicely formatted string via markdown/MathJax notation
>
> -   If your output table spans across two pages, try inserting a page break in the markdown just before your code chunk.

\newpage

```{r Data Summary, warning=FALSE,error=FALSE,echo=FALSE}
#compute mean PM2.5 concentration for each county, sorted by value
summary_PM25 <- epa_data %>% 
  drop_na(PM2.5) %>% 
  group_by(COUNTY) %>% 
  summarize(mean_PM2.5 = mean(PM2.5)) %>% 
  arrange(mean_PM2.5)
#rename column names (question: another approach?)
colnames(summary_PM25)<-c("Country","$\\mu$g/m$^3$")


#compute mean ozone concentration for each county, sorted by value
summary_Ozone <- epa_data %>% 
  drop_na(Ozone) %>% 
  group_by(COUNTY) %>% 
  summarize(
    mean_Ozone = mean(Ozone)) %>% 
  arrange(mean_Ozone)
#rename column names 
colnames(summary_Ozone)<-c("Country","$\\mu$g/m$^3$")

#use the `kable` function to produce the 2 tables in markdown format
knitr::kable(summary_PM25,caption="Mean Particulates (2.5mm)")
knitr::kable(summary_Ozone,caption="Mean Ozone")

#question: adjust the size of the figures 
```

------------------------------------------------------------------------

## Task 3: Plots

Create two separate code chunks that create boxplots of the distribution of Ozone levels by month using, one for only records collected in 2018 and one for records in 2019. Customize the chunk options such that the final figures are displayed but not the code used to generate the figures. In addition, align the plots on the left side of the page and set the figure heights so both plots fit on the same page with minimal space remaining. Lastly, add a `fig.cap` chunk option to add a caption (title) to your plot that will display underneath the figure.

```{r Ozone 2018, warning=FALSE,error=FALSE,echo=FALSE, fig.cap="Figure. Monthly Ozone in 2018",out.height="40%"}
#filter year
df_Ozone_2018<-epa_data %>%
  filter(Year==2018) 



#create a boxplot by month
Ozone_2018<-ggplot(df_Ozone_2018)+
  geom_boxplot(aes(x=factor(Month,levels=1:12,labels=month.abb),y=Ozone))+
  labs(x="Month",y=expression(Concentration~(mu*g/m^3)))+
  ggtitle("Recorded Ozone by Month (2018)")

Ozone_2018

```

```{r Ozone 2019, warning=FALSE,error=FALSE,echo=FALSE, fig.cap="Figure. Monthly Ozone in 2019",out.height="40%" }
#filter year
df_Ozone_2019<-epa_data %>%
  filter(Year==2019) 

#create a boxplot by month

Ozone_2019<-ggplot(df_Ozone_2019)+
  geom_boxplot(aes(x=factor(Month,levels=1:12,labels=month.abb),y=Ozone))+
  labs(x="Month",y=expression(Concentration~(mu*g/m^3)))+
  ggtitle("Recorded Ozone by Month (2019)")


Ozone_2019
```

------------------------------------------------------------------------

\newpage

## Task 4: Knit and submit.

Add a table of contents to your document and knit to a PDF. Submit your PDF to Sakai, but also be sure to commit and push your Rmd file used to create this knit document to GitHub. In the section below, add a link to your GitHub repository.

## Git Repository

[Link of Git Repositiry](https://github.com/inaliao/EDE_Fall2023/tree/main/Assignments)
