---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Ina Liao"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.
```{r Install Packages,message=FALSE,error=FALSE,results = FALSE}
#install.packages("tidyverse")
#install.packages("agricolae")
#install.packages("here")
#install.packages("ggplot2")
#install.packages("cowplot")
#install.packages("ggridges")
#install.packages("viridis")
#install.packages("RColorBrewe")
#install.packages("colormap")
#install.packages("ggthemes")
#install.packages("dplyr")

library(tidyverse)
library(agricolae)   
library(here)
library(ggplot2)
library(cowplot)
library(ggridges)
library(viridis)
library(RColorBrewer)
library(colormap)
library(ggthemes)
library(dplyr)
here()
```

```{r Import Data,warning=FALSE,error=FALSE}
df_NTL_LTER<-read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"))
#date format
df_NTL_LTER$sampledate<-as.Date(df_NTL_LTER$sampledate,format="%m/%d/%y")
```
2. Build a ggplot theme and set it as your default theme.

```{r Build Theme,warning=FALSE,error=FALSE}
plot_theme<-theme_calc()+
  theme(
    #plot background
    plot.background=element_rect(color="gray"),
    
    #plot title
    plot.title=element_text(color="black",hjust=0.5,vjust=1),
    
    #axis labels
    axis.text=element_text(color="black"),
    
    #legend
    legend.key=element_rect(color="white"),
    legend.background = element_rect(color="white"),
    legend.position="right",
    
    #gridlines 
    panel.grid.major=element_line("white"),
    axis.ticks=element_line(color="white")
  )
theme_set(plot_theme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature recorded during July does not change with depth across all lakes.
Ha: The mean lake temperature recorded during July does change with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)
 
```{r Wrangle Data, warning=FALSE}
df_July<-df_NTL_LTER %>%
  mutate(month=format(sampledate,"%B"))%>%
  filter(month=="July") %>%
  select(lakename,year4,daynum,depth,temperature_C) %>%
  na.omit()
```

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r Scatterplot,warning=FALSE,error=FALSE,fig.cap="Temperature by depth" }
plot_scatter<-ggplot(df_July,aes(x=temperature_C,y=depth,color=temperature_C))+
  geom_point()+
  xlim(0,35)+
  scale_y_reverse()+
  scale_color_gradient(low = "blue", high = "orange")+
  geom_smooth(method="lm",color="black")+
  labs(x="temperature (C)",y="depth (m)")
print(plot_scatter)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: As the depth of the water increases, the temperature decreases, which means that there might be a negative linear relationship between depth and temperature. 

7. Perform a linear regression to test the relationship and display the results

```{r Linear Regression, warning=FALSE}
tem_dep<-lm(df_July$temperature_C~df_July$depth)
summary(tem_dep)
```

8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The residual of the model ranges from -9.5 to 13.6. The slope of the regression line is -1.94 (df=9726), meaning there is a negative relationship between temperature and depth level. The coefficient is statistically significant different from 0 (p-value<0.001), meaning that we have enough evidence to reject that that mean lake temperature recorded during July does not change with depth across all lakes. The r-sqaure value is 0.738, indicating that a unit increase in depth is associated with a 0.738 unit increase in the predicted value of temperature. 

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 

9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

```{r temperature.model, warning=FALSE}
#9 AIC test
test_AIC<-lm(df_July$temperature_C~df_July$year4 + df_July$daynum + df_July$depth)
step(test_AIC)

#10 linear model
model_linear<-lm(df_July$temperature_C~ df_July$daynum + df_July$depth)
summary(model_linear)
```

> Answer: The AIC test suggests us to use "daynum" and "depth" as explanatory variables in the multiple regreesion to predict temperature. The model indicate that a unit increase in daynum and depth is associated with a 0.741 unit increase in the predicted temperature value, which has higher prediction than the previous model (r-square: 0.741>0.728).

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

Ho: the mean temperatures in each lake is the same in the month of July
Ha: the mean temperatures in each lake is not the same in the month of July

```{r anova.model, warning=FALSE}
# wrangle data 
df_tem_lake<-df_July %>%
  group_by(lakename,year4,daynum) %>%
  summarise(temperature_C)

# one-way ANOVA
tem_ANOVA<-aov(data=df_tem_lake,temperature_C~lakename)
summary(tem_ANOVA)

# format ANOVA as linear model
tem_ANOVA_lm<-lm(data=df_tem_lake,temperature_C~lakename)
summary(tem_ANOVA_lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The coefficient is statistically significant different from 0 (p-value<0.001), meaning that we have enough evidence to reject that the temperatures in each lake is the same in the month of July. 

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, warning=FALSE,fig.cap="Temperature by depth in each lake"}
#14.
plot_tem_lake<-ggplot(df_July,aes(x=temperature_C,y=depth,color=lakename))+
  geom_point(alpha=0.5)+
  xlim(0,35)+
  scale_y_reverse()+
  labs(x="temperature (C)",y="depth (m)",color="Lake Name")+
  theme(legend.position="right",
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 10))
print(plot_tem_lake)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test, warning=FALSE, fig.cap="Mean temperature in each lake"}
#15
# HSD test
TukeyHSD(tem_ANOVA)
test_HSD<-HSD.test(tem_ANOVA,"lakename",group = TRUE)
test_HSD

# graph the results
plot_HSD<-ggplot(df_tem_lake, aes(x = lakename, y = temperature_C)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("a", "b", "c", "d", "de", "e", 
                         "e", "f", "f")) +
  labs(x = "Lakename", y = "Temperature (C)") 
print(plot_HSD)
```

>All lakes are grouped with at least another lake; thus, no lake has a different means than others. 


16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Statistically, Peter Lake has the same mean temperature as Paul Lake (p>0.001). 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can look into the interaction effect of Pater Lake and Paul Lake on the mean temperature to understand if the effect of Pater Lake temperature and Paul Lake temperature depends on each other. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

Ho: the mean temperature in Crampton Lake and Ward Lake is the same in the month of July 
Ha: the mean temperature in Crampton Lake and Ward Lake is not the same in the month of July 


```{r t.test, warning=FALSE}
# filter data
df_July_sub<-df_July %>%
  filter(lakename=="Crampton Lake" | lakename=="Ward Lake") 

# two-sample t-test
test_twosample<-t.test(df_July_sub$temperature_C~df_July_sub$lakename)
test_twosample
```

>Answer: We do not have enough evidence to reject that the mean temperature in Crampton Lake and Word Lake are the same (p-value>0.1, df=200), which matches the answer in question 16. 
